---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggmHMM

<!-- badges: start -->
<!-- badges: end -->

The goal of ggmHMM is to visualize the output of Multilevel Hidden Markov Models (MHMMs) using the [ggplot2](https://ggplot2.tidyverse.org/) package. It is meant to be used alongside the [mHMMbayes](https://github.com/emmekeaarts/mHMMbayes/) package. Note that the package is **a work in progress**. Please report any issues using the [issues](https://github.com/pepijnvink/ggmHMM/issues) tab on GitHub.

## Installation

You can install the development version of ggmHMM from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("pepijnvink/ggmHMM")
```

## Example

To show an example of how the package can work, we first load the ggmHMM package and the mHMMbayes package, and simulate some data, as well as run a model.

```{r example-notrun, eval = FALSE}
library(ggmHMM)
library(mHMMbayes)
library(ggplot2)
## basic example code
# simulating multivariate continuous data
n_t     <- 100
n       <- 10
m       <- 3
n_dep   <- 2

gamma   <- matrix(c(0.8, 0.1, 0.1,
                    0.2, 0.7, 0.1,
                    0.2, 0.2, 0.6), ncol = m, byrow = TRUE)

emiss_distr <- list(matrix(c( 50, 10,
                              100, 10,
                              150, 10), nrow = m, byrow = TRUE),
                    matrix(c(5, 2,
                             10, 5,
                             20, 3), nrow = m, byrow = TRUE))
set.seed(42)
data_cont <- sim_mHMM(n_t = n_t, n = n, data_distr = 'continuous',
                      gen = list(m = m, n_dep = n_dep),
                      gamma = gamma, emiss_distr = emiss_distr,
                      var_gamma = .1, var_emiss = c(5^2, 0.2^2))

# Specify hyper-prior for the continuous emission distribution
manual_prior_emiss <- prior_emiss_cont(
                        gen = list(m = m, n_dep = n_dep),
                        emiss_mu0 = list(matrix(c(30, 70, 170), nrow = 1),
                                         matrix(c(7, 8, 18), nrow = 1)),
                        emiss_K0 = list(1, 1),
                        emiss_V =  list(rep(5^2, m), rep(0.5^2, m)),
                        emiss_nu = list(1, 1),
                        emiss_a0 = list(rep(1.5, m), rep(1, m)),
                        emiss_b0 = list(rep(20, m), rep(4, m)))

# Run the model on the simulated data:
# Note that for reasons of running time, J is set at a ridiculous low value.
# One would typically use a number of iterations J of at least 1000,
# and a burn_in of 200.
out_3st_cont <- mHMM(s_data = data_cont$obs,
                         data_distr = 'continuous',
                         gen = list(m = m, n_dep = n_dep),
                         start_val = c(list(gamma), emiss_distr),
                         emiss_hyp_prior = manual_prior_emiss,
                         mcmc = list(J = 100, burn_in = 5))
```

```{r example, include = FALSE}
library(ggmHMM)
library(mHMMbayes)
library(ggplot2)
mhmm_cont <- readRDS("tests/testthat/fixtures/mhmm_cont.rds")
```

We can now, for example, create a plot to visualize the emission means, using `plot_emiss()`, and include the subject-specific effects. We can then use functions from the [ggplot2](https://ggplot2.tidyverse.org/) package to further customize the plot, such as adding a title with `ggtitle()`.

```{r example-plot-emiss}
plot_emiss(mhmm_cont, type = "bar", subject_effects = TRUE, line = TRUE) +
  ggtitle("Emission means with random effects")
```
