---
title: "ggmHMM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggmHMM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `ggmHMM` package provides wrapper functions used to visualize Bayesian Multilevel Hidden Markov Models, using the `ggplot2` package. Most functions take an object that is built using the `mHMMbayes` package. They output a 'ggplot' object, which can be edited by a user using functions from the `ggplot2` package. This vignette will not go into details of the Bayesian Multilevel Hidden Markov Model, neither into fitten a model; for that, see the documentation of the `mHMMbayes` package.

# Set-up
First, we load the required packages:
```{r}
library(ggmHMM)
library(ggplot2)
library(mHMMbayes)
```

For our illustrations, we will use the `mHMMbayes::nonverbal` dataset. This is a dataset describing the communication dynamics of 10 patient-therapist dyads during a therapy session.

```{r}
nonverbal <- data.frame(nonverbal)
```

The dataset contains 5 columns:
* `id`: id variable of patient - therapist couple to distinguish which observation belongs to which couple.
* `p_verbalizing`: verbalizing behavior of the patient, consisting of 1 = not verbalizing, 2 = verbalizing, 3 = back channeling.
* `p_looking`: looking behavior of the patient, consisting of 1 = not looking at therapist, 2 = looking at therapist.
* `t_verbalizing`: verbalizing behavior of the therapist, consisting of 1 = not verbalizing, 2 = verbalizing, 3 = back channeling.
* `t_looking`: looking behavior of the therapist, consisting of 1 = not looking at patient, 2 = looking at patient.
The top 6 rows of the dataset are provided below. 

```{r}
head(nonverbal)
```

## Fit model
We will fit a simple 2-state model using the `mHMMbayes` package:

```{r settings and load models, include = FALSE}
nonverbal_out_2st <- readRDS("out_nonverbal_2st_1000iter.rds")
nonverbal_out_2st_2 <- readRDS("out_nonverbal_2st_1000iter_2.rds")
out_3st_cont <- readRDS("out_cont_3st.rds")
```

```{r show specifying 2 state model, eval = FALSE}
# specifying general model properties:
m <- 2
n_dep <- 4
q_emiss <- c(3, 2, 3, 2)

# specifying starting values
start_TM <- diag(.8, m)
start_TM[lower.tri(start_TM) | upper.tri(start_TM)] <- .2
start_EM <- list(matrix(c(0.05, 0.90, 0.05, 
                          0.90, 0.05, 0.05), byrow = TRUE,
                        nrow = m, ncol = q_emiss[1]), # vocalizing patient
                 matrix(c(0.1, 0.9, 
                          0.1, 0.9), byrow = TRUE, nrow = m,
                        ncol = q_emiss[2]), # looking patient
                 matrix(c(0.90, 0.05, 0.05, 
                          0.05, 0.90, 0.05), byrow = TRUE,
                        nrow = m, ncol = q_emiss[3]), # vocalizing therapist
                 matrix(c(0.1, 0.9, 
                          0.1, 0.9), byrow = TRUE, nrow = m,
                        ncol = q_emiss[4])) # looking therapist
# Run a model without covariate(s) and default priors:
set.seed(42)
nonverbal_out_2st <- mHMM(s_data = nonverbal, 
                gen = list(m = m, n_dep = n_dep, q_emiss = q_emiss), 
                start_val = c(list(start_TM), start_EM),
                mcmc = list(J = 1000, burn_in = 200))

# Run a second chain
## specifying starting values for seconf chain
start_TM_2 <- diag(.7, m)
start_TM_2[lower.tri(start_TM_2) | upper.tri(start_TM_2)] <- .3
start_EM_2 <- list(matrix(c(0.1, 0.80, 0.1, 
                            0.8, 0.1, 0.1), byrow = TRUE,
                        nrow = m, ncol = q_emiss[1]), # vocalizing patient
                 matrix(c(0.2, 0.8, 
                          0.2, 0.8), byrow = TRUE, nrow = m,
                        ncol = q_emiss[2]), # looking patient
                 matrix(c(0.80, 0.1, 0.1, 
                          0.1, 0.8, 0.1), byrow = TRUE,
                        nrow = m, ncol = q_emiss[3]), # vocalizing therapist
                 matrix(c(0.2, 0.8, 
                          0.2, 0.8), byrow = TRUE, nrow = m,
                        ncol = q_emiss[4])) # looking therapist
set.seed(123)
nonverbal_out_2st_2 <- mHMM(s_data = nonverbal, 
                gen = list(m = m, n_dep = n_dep, q_emiss = q_emiss), 
                start_val = c(list(start_TM_2), start_EM_2),
                mcmc = list(J = 1000, burn_in = 200))
```



We will also simulate continuous data and fit a mHMM on it
```{r show specifying continuous model, eval = FALSE}
# Continuous Dependent Variables
# simulating multivariate continuous data
n_t     <- 100
n       <- 10
m       <- 3
n_dep   <- 2

gamma   <- matrix(c(0.8, 0.1, 0.1,
                    0.2, 0.7, 0.1,
                    0.2, 0.2, 0.6), ncol = m, byrow = TRUE)

emiss_distr <- list(matrix(c( 50, 10,
                              100, 10,
                              150, 10), nrow = m, byrow = TRUE),
                    matrix(c(5, 2,
                             10, 5,
                             20, 3), nrow = m, byrow = TRUE))
set.seed(42)
data_cont <- sim_mHMM(n_t = n_t, n = n, data_distr = 'continuous',
                      gen = list(m = m, n_dep = n_dep),
                      gamma = gamma, emiss_distr = emiss_distr,
                      var_gamma = .1, var_emiss = c(5^2, 0.2^2))

# Specify hyper-prior for the continuous emission distribution
manual_prior_emiss <- prior_emiss_cont(
  gen = list(m = m, n_dep = n_dep),
  emiss_mu0 = list(matrix(c(30, 70, 170), nrow = 1),
                   matrix(c(7, 8, 18), nrow = 1)),
  emiss_K0 = list(1, 1),
  emiss_V =  list(rep(5^2, m), rep(0.5^2, m)),
  emiss_nu = list(1, 1),
  emiss_a0 = list(rep(1.5, m), rep(1, m)),
  emiss_b0 = list(rep(20, m), rep(4, m)))

# Run the model on the simulated data:
# Note that for reasons of running time, J is set at a ridiculous low value.
# One would typically use a number of iterations J of at least 1000,
# and a burn_in of 200.
out_3st_cont <- mHMM(s_data = data_cont$obs,
                     data_distr = 'continuous',
                     gen = list(m = m, n_dep = n_dep),
                     start_val = c(list(gamma), emiss_distr),
                     emiss_hyp_prior = manual_prior_emiss,
                     mcmc = list(J = 1000, burn_in = 200))
```


# Evaluate Convergence
We will start by evaluating convergence at the group level using trace plots. We can do this using `plot_convergence()`. We will start with trace plots for gamma (i.e. the transition probabilities):

We first look at plotting the intercepts over iterations, corresponding to the transition probabilities:

```{r plot gamma-group intercept}
plot_convergence(
  model = nonverbal_out_2st, # the model to plot
  component = "gamma", # plot transition probabilities
  level = "group", # group-level parameters
)
```

We can also choose to plot it on the probability scale, rather than intercept scale, by including `prob = TRUE`:

```{r plot gamma-group prob}
plot_convergence(
  model = nonverbal_out_2st, # the model to plot
  component = "gamma", # plot transition probabilities
  level = "group", # group-level parameters
  prob = TRUE # plot at probability level (rather than intercept-level)
)
```

We can also plot the group-level variances instead of probabilities. To do this, we specify `param = "var"`:
```{r plot gamma-var prob}
plot_convergence(
  model = nonverbal_out_2st, # the model to plot
  component = "gamma", # plot transition probabilities
  param = "var", # plot group_level variances
  level = "group", # group-level parameters
)
```

If, rather than transition probabilities, you want to obtain trace plots for the emission distribution, you specify `component = "emiss"` and a variable to plot:

```{r plot emiss-group}
plot_convergence(
  model = nonverbal_out_2st, # the model to plot
  component = "emiss", # plot transition probabilities
  vrb = "p_looking", # variable to plot
  level = "group", # group-level parameters
  prob = TRUE # specify FALSE to plot on intercept scale instead
)
```

If, instead of group-level you want to plot subject-level trace plots, you simply specify `level = "subject"` and provide a subject ID to plot using the argument `ID`. So if we want to plot the emission probabilities of the variable `p_looking` for subject `8`, we specify:
```{r plot emiss-subj}
plot_convergence(
  model = nonverbal_out_2st, # the model to plot
  component = "emiss", # plot transition probabilities
  vrb = "p_looking", # variable to plot
  level = "subject", # subject-level parameters
  ID = 8, # subject to plot for
  prob = TRUE # specify FALSE to plot on intercept scale instead
)
```

The default in `plot_convergence()` is to label the categories and states using numbers. However it is possible to provide state labels using the `state_labels` argument (and for categorical emissions, category labels using the `cat_labels` argument). These can be specified as follows:
```{r plot emiss-subj labeled}
plot_convergence(
  model = nonverbal_out_2st, # the model to plot
  component = "emiss", # plot transition probabilities
  vrb = "p_looking", # variable to plot
  level = "subject", # subject-level parameters
  ID = 8, # subject to plot for
  state_labels = c("Patient Speaking State", "Patient Listening State"),
  cat_labels = c("No Looking", "Looking"),
  prob = TRUE # specify FALSE to plot on intercept scale instead
)
```

It is also possible to plot multiple chains simultaneously. This can be done by probiding a list of `mHMMbayes::mmHMM` objects:
```{r plot emiss-subj labeled twochains}
plot_convergence(
  model = list(nonverbal_out_2st,
               nonverbal_out_2st_2), # list of models to plot
  component = "emiss", # plot transition probabilities
  vrb = "p_looking", # variable to plot
  level = "subject", # subject-level parameters
  ID = 8, # subject to plot for
  state_labels = c("Patient Speaking State", "Patient Listening State"),
  cat_labels = c("No Looking", "Looking"),
  prob = TRUE # specify FALSE to plot on intercept scale instead
)
```

# Plotting Continuous Emission Distributions
`plot_emiss()` can be used to plot continuous emission distributions using a bar plot or a boxplot. To illustrate this, we will use our simulated continuous data. You can specify plotting a bar plot of emissions as follows:
```{r}
plot_emiss(
  model = out_3st_cont, # specify model to plot
  type = "bar", # obtain a bar chart
  individual = FALSE # do not overlay individual-level points
)
```

When we choose a bar plot, we can also choose to overlay the subject-level estimates by specifying `individual = TRUE`. We can then also specify the transparency of the points representing the subject-level means using the `alpha` argument:
```{r}
plot_emiss(
  model = out_3st_cont, # specify model to plot
  type = "bar", # obtain a bar chart
  individual = TRUE, # do not overlay individual-level points
  alpha = 0.5 # specify transparency of points
)
```

Another option is to plot boxplots of the subject-level emission means. This can be achieved by instead specifying `type = "boxplot"`:
```{r}
plot_emiss(
  model = out_3st_cont, # specify model to plot
  type = "boxplot", # obtain a bar chart
  state_labels = c("red", "orange", "green"), # we can give custom labels to the states
)
```

# Plot transition probabilities
`plot_gamma()` can be used to obtain a heat plot for the transition probabilities. To obtain a heat plot for the group-level transaition probabilities for the `nonverbal` dataset, specify:
```{r heatplot group-mHMM}
plot_gamma(
  model = nonverbal_out_2st, # mHMM model
  level = "group", # plot group level
  digits = 3 # number of digits to round to
)
```

To instead plot subject-level transition probabilities, specify the following:
```{r}
plot_gamma(
  model = nonverbal_out_2st, # mHMM model
  level = "subject", #plot subject level
  facet = TRUE
)
```

To only plot transition probabilities for a subset of individuals, you can provide an integer (or integer vector) to the argument `ID`, specifying the subject indices:
```{r}
plot_gamma(
  model = nonverbal_out_2st, # mHMM model
  level = "subject", #plot subject level
  ID = 1:5, # only plot for subjects 1 to 5
  facet = TRUE
)
```

The default for subject level heat plots is to show separate plots for each subject. Alternatively, you can create a single heat plot, with the subjects as the rows and the transition probabilities as the columns, by specifying `facet = FALSE`:
```{r}
plot_gamma(
  model = nonverbal_out_2st, # mHMM model
  level = "subject", #plot subject level
  ID = 1:5, # only plot for subjects 1 to 5
  facet = FALSE # do not facet
)
```

Alternatively, instead of a `mHMM:mHMM` object and a level specification, a, object created using [mHMMbayes:obtain_gamma()] may be provided:
```{r}
## Obtain group-level transition probabilities
gamma_group <- obtain_gamma(
  object = nonverbal_out_2st,
  level = "group"
)
## Plot group-level transition probabilities
plot_gamma(
  model = gamma_group, # transition probabilities. 'level' argument is now redundant
)
```

The same can be done for subject-level transition probabilities:

```{r}
## Obtain group-level transition probabilities
gamma_subj <- obtain_gamma(
  object = nonverbal_out_2st,
  level = "subject"
)
## Plot group-level transition probabilities
plot_gamma(
  model = gamma_subj, # transition probabilities. 'level' argument is now redundant
)
```

## Plot posterior distributions
In addition to plotting the posterior estimates and trace plots for the posterior distributions, it is also possible to visualize density plots of posterior distributions. `plot_posterior()` can be used to create separate density plots for all individuals, overlayn with the group-level posterior distributions. `plot_posterior(object)` can be used as a replacement for the `plot` method (i.e. `plot(object)`) in the `mHMMbayes` package. It works as follows for plotting transition probabilities:
```{r}
# Plot transition probabilities
plot_posterior(
  model = nonverbal_out_2st, # model to plot
  component = "gamma", # plot transition distributions
  state_labels = c("Patient Speaking State", "Patient Listening State") # again, optional
)
```

It can also be used to plot emission distributions for categorical or continuous data. We can do this by specifying `component = "emiss", and providing a variable to plot. For categorical data, it plots the emission probabilities at each level of the variable:
```{r}
plot_posterior(
  model = nonverbal_out_2st,
  component = "emiss",
  vrb = "p_looking", # specify variable to plot
  cat_label = c("No Looking", "Looking") # again, optional
)
```

For continuous data, `plot_posterior()` will plot both the means and the standard deviations in a single plot. Because these standard deviations are fixed over persons, these will not show subject-level standard deviations. Now, we will also decrease the transparancy of the subject-level densities by specifying a higher alpha (i.e. `alpha = 0.5` rather than the default `alpha = 0.1`).
```{r}
plot_posterior(
  model = out_3st_cont,
  component = "emiss",
  vrb = "observation 1",
  alpha = 0.5 # less transparency
)
```

# Plot inferred states obtained using the viterbi algorithm
The `mHMMbayes` package contains the function [mHMMbayes::vit_mHMM()], which can be used to obtain inferred states for each subject, based on their maximum a posteriori estimates for the emission distibutions and transition probabilities. `plot_viterbi()` takes a data frame obtained using [mHMMbayes::vit_mHMM()] and plots these states for each specified person. It works as such:
```{r}
# obtain inferred states
states <- vit_mHMM(object = nonverbal_out_2st, s_data = nonverbal)
plot_viterbi(
  states = states # dataframe with states
)
```

As shown above, the default is to plot infered states for all individuals. In the plot below, the states for only a subset of individuals is plotted. In addition, we specify a vector of colors to assign to each state, as well as names to assign to each state.
```{r}
plot_viterbi(
  states = states, # dataframe with states
  colors = c("blue", "orange"), # specify colors to use
  ID = 1:5, # vector specifying subjects to plot. integer for a single subject is also possible
  state_labels = c("Patient Speaking State", "Patient Listening State")
)
```

# Editing plot
Each plotting function from `ggmHMM` returns an object of type `ggplot`. This means that functions from the `ggplot2` package can be used to customize plots, such as customizing themes.
The default theme for plots returned by functions in the package is the `classic` theme, for example:
```{r}
# Plot transition probabilities
posterior <- plot_convergence(
  model = list(nonverbal_out_2st, nonverbal_out_2st_2), # the model to plot
  component = "gamma", # plot transition probabilities
  level = "group", # group-level parameters
  prob = TRUE # plot at probability level (rather than intercept-level)
)
posterior
```

If instead you want to use another theme, such as the `minimal` theme, this can be achieved by simply calling `theme_minimal()` as you usually would:
```{r}
posterior + theme_minimal()
```

Similarly is also possible to add a title and/or customize the axis labels, edit line colors, and more!
```{r}
posterior +
  theme_dark() + # set new theme
  labs(title = "Trace Plot for Group-Level Transition Probabilities", # add title
       x = "Iteration", # change x axis title
       y = "Transition Probability") +# change y axis title
  scale_color_manual(values=c("green", "orange")) # change colors for each chain
```

